parallel
--------

The `parallel` package is derived from the `snow` and `multicore` packages. Many of the functions from the packages can be used with minor changes.

The `parallel` package integrates with the L'Ecuyer-CMRG random number generator (RNG) in R 2.14.0. The seed of this generator can be advanced a fixed number of steps making it a useful `parallel` RNG. `multicore` derived functions now have true RNG support.

### How It Works

The `parallel` package can run on Posix-based multicore-based systems using functions such as `mclapply()` and `mcparallel()` derived from the `multicore` package. It can also be used on a PSOCK cluster and functions such as `parLapply()` and `clusterApplyLB` will execute on multicore Windows systems and Linux clusters. Since parallel can be used on cluster objects created by `snow`, the MPI transport can be used.

### Working with It

For Posix-based systems, functions like `mclapply` can be used. However, used `detectCores()` rather than the `mc.cores` argument.

```{r}
library(parallel)
library(MASS)
RNGkind("L'Ecuyer-CMRG")
mc.cores <- detectCores()
results <- mclapply(rep(25, 4), function(nstart) kmeans(Boston, 4, nstart=nstart),
  mc.cores=mc.cores)
i <- sapply(results, function(result) result$tot.withinss)
result <- results[[which.min(i)]]
result
```

For Windows, use the snow derived APIs in `parallel`. For example:
```{r}
# options(mc.cores=detectCores())
# cl <- makeCluster(4)
cl <- makeCluster(detectCores())
clusterSetRNGStream(cl)
clusterEvalQ(cl, library(MASS))
results <- clusterApply(cl, rep(25, 4), function(nstart) kmeans(Boston, 4,
  nstart=nstart))
i <- sapply(results, function(result) result$tot.withinss)
result <- results[[which.min(i)]]
result
stopCluster(cl)
```

### Creating Clusters with makeCluster

You cannot use `multicore` derived functions such as `mclapply()` and `pvec()` on Windows or Linux clusters. Instead use `snow` derived functions such as `parLapply()` or `clusterApplyLB()`.

The `parallel` package comes with two transports: `PSOCK` and `FORK`. `PSOCK` is a streamlined verion of snow's `SOCK` and communicates between the master and the workers using socket connections. In the `makeCluster()` function, `PSOCK` is the default type.

The `FORK` transport starts workers using the `mcfork()` function and communicates between the master and the workers using socket connections. `mcfork()` builds on `fork()` and thus a Linux cluster is not supported. Since it is Posix based, Windows is not supported. 

```{r}
library(parallel)
cl <- makeCluster(4)
stopCluster(cl)
cl <- makeCluster(detectCores())
stopCluster(cl)
cl <- makeCluster(4, type="FORK")
stopCluster(cl)
```

`FORK` workers inherit the data and the environment of the master. However, a `FORK` cluster is persistent. Thus, you should create a `FORK` cluster before calling `parLapply`.

Code for Posix-based systems and Windows:
```{r}
library(parallel)
# type <- if (exists("mcfork", mode="function")) "FORK" else "PSOCK"
cores <- getOption("mc.cores", detectCores())
cl <- makeForkCluster(cores)
# cl <- makeCluster(cores, type=type)
results <- parLapply(cl, 1:10, sqrt)
results
stopCluster(cl)
```

### Parallel Random Number Generation

The parallel package supports parallel random number generation using ideas from the `rlecuyer` package.

`RNGkind()` sets the random number generator. Leave `mc.set.seed` set to `TRUE`. For `multicore` derived functions, each worker gets a new random number stream.

```{r}
library(parallel)
RNGkind("L'Ecuyer-CMRG")
mclapply(1:2, function(i) rnorm(1))
```

If `mc.reset.stream()` is called, the parallel random number generator is reinitialized using the master seed.

```{r}
RNGkind("L'Ecuyer-CMRG")
set.seed(7777442)
mc.reset.stream()
unlist(mclapply(1:2, function(i) rnorm(1)))
set.seed(7777442)
mc.reset.stream()
unlist(mclapply(1:2, function(i) rnorm(1)))
```

To use parallel RNG in snow-derived functions, use `clusterSetRNGStream()`. The seed is an argument to this function.

```{r}
library(parallel)
cl <- makeCluster(4, type = "FORK")
clusterSetRNGStream(cl, 7777442)
unlist(clusterEvalQ(cl, rnorm(1)))
clusterSetRNGStream(cl, 7777442)
unlist(clusterEvalQ(cl, rnorm(1)))
cl
stopCluster(cl)
```

The `parallel` package includes utilities for advancing the seed: `nextRNGStream()` and `nextRNGSubStream()`.

```{r}
.Random.seed <<- nextRNGSubStream(.Random.seed)
```

## Summary of Differences

See text.



