Assignment 4: Streams
--------------------------------

Generate bivariate normal streaming data with different means represening three clusters.

The objective of this exercise is to gererate, expore, and evaluate streaming data. You need to explain the code in each item and discuss the output. Refer to the help files for specific functions.

1. Generate bivariate normal streaming data with different means represening three clusters. Explain how and what is generated, i.e., explain `dsd`.

```{r}
library("stream")
set.seed(1002)
dsd <- DSD_Gaussians(k=3, d=2, mu=rbind(c(1.5,1.3),c(1,1),c(1.2,1)))
dsd
```

The function `DSD_Gaussians()` creates an eponymous object (which
inherits from `DSD`, the Data Stream Data Generator class) containing
parameters used to generate observations from the simulated data
stream.  The stream simulates three clusters, each made up of
a bivariate normal mixture model with means defined the mu matrix and
randomly generated covariance matrices.


2. Perform a threshold nearest neighbor micro-clustering. Explain what is generated at every step. Do the evaluations measures provide good fits?

```{r}
tnn <- DSC_tNN(r=.1)
tnn
update(tnn, dsd, n=500, verbose=FALSE)
tnn
head(get_centers(tnn))
evaluate(tnn, dsd, measure = c("purity", "crand"), n = 500)
```

First, we initialize tNN object with a threshold of 0.1, we see that
by default no clusters have been formed since no data has been processed.

After this, we update the model with 500 observations from our data
stream silently and examine the results.  We find that two of our
macro-clusters were identified along with twelve micro-clusters.

We see three clusters centered near out first true cluster with means
(1.5, 1.3) and three somewhat near our second true cluster with means
(1, 1).  The  third true cluster with means (1.2, 1) looks to have
been lost in the noise.

Our purity measure is fairly high, but the corrected Rand index (a
measure of the accuracy) tells us we're more dissimilar to our true
clusters than similar.

3. Plot the micro-clusters. Interpret the plot.

```{r}
plot(tnn, dsd)
```

We can see our twelve micro-clusters plotted on a scatterplot of our
data. The micro-clusters are represented as circles whose radii
represent the strength of the cluster.

Note that the micro-clusters 

4. Perform and plot a $k$-means macro cluster. Explain what is generated at every step. Interpret the plot.

```{r}
kmeans <- DSC_Kmeans(k=3)
recluster(kmeans, tnn)
plot(kmeans, dsd, type = "both")
```

<!-- Put your comments here -->

5. Do the macro evaluations measures provide good fits? Contrast with 2.

```{r}
evaluate(kmeans, dsd, measure = c("purity", "crand"), n = 500)
```

<!-- Put your comments here -->






