Assignment 4: Streams
--------------------------------
### Doug Raffle (dcraffle@mix.wvu.edu)

Generate bivariate normal streaming data with different means represening three clusters.

The objective of this exercise is to gererate, expore, and evaluate streaming data. You need to explain the code in each item and discuss the output. Refer to the help files for specific functions.

1. Generate bivariate normal streaming data with different means represening three clusters. Explain how and what is generated, i.e., explain `dsd`.

```{r}
library("stream")
set.seed(1002)
dsd <- DSD_Gaussians(k=3, d=2, mu=rbind(c(1.5,1.3),c(1,1),c(1.2,1)))
dsd
```

The function `DSD_Gaussians()` creates an eponymous object (which
inherits from `DSD`, the Data Stream Data Generator class) containing
parameters used to generate observations from the simulated data
stream.  The stream simulates three clusters, each made up of
a bivariate normal mixture model with means defined the mu matrix and
randomly generated covariance matrices.


2. Perform a threshold nearest neighbor micro-clustering. Explain what is generated at every step. Do the evaluations measures provide good fits?

```{r}
tnn <- DSC_tNN(r=.1)
tnn
update(tnn, dsd, n=500, verbose=FALSE)
tnn
head(get_centers(tnn))
evaluate(tnn, dsd, measure = c("purity", "crand"), n = 500)
```

First, we initialize tNN object with a threshold of 0.1, we see that
by default no clusters have been formed since no data has been processed.

After this, we update the model with 500 observations from our data
stream silently and examine the results.  We find that two of our
macro-clusters were identified along with twelve micro-clusters.

We see that, of the first six clusters, three clusters centered near
our first true cluster with means 
(1.5, 1.3) and three somewhat near our second true cluster with means
(1, 1). 

Our purity measure is fairly high, so points from the true clusters
tend to end up together.  The corrected Rand index (a
measure of the accuracy of the structure) tells us we're more dissimilar to our true
clusters than similar, this is likely the result of overfitting --
we're finding more clusters than should exist.

3. Plot the micro-clusters. Interpret the plot.

```{r}
plot(tnn, dsd)
```

The plot shows the cluster centroids represented as circles overlayed
on a scatterplot of our distributions.  The relative strength of the
clusters is represented by their radii.

Most of the microclusters are falling near where they should. The true
clusters centered at (1, 1)  and (1.5, 1.2) have a lot of variance, so the clusters
falling in this area are less compact. Contrast this with the
relatively high covariance of the cluster centered at (1.2, 1) whose
clusters are much stronger.


4. Perform and plot a $k$-means macro cluster. Explain what is generated at every step. Interpret the plot.

```{r}
kmeans <- DSC_Kmeans(k=3)
recluster(kmeans, tnn)
plot(kmeans, dsd, type = "both")
```

In the first expression, we initialize a $k$-means cluster object
primed to cluster with three centroids.

This clustering scheme is then added to our previously constructed
threshold cluster object, which we then visualize.

We see that the centroids of the $k$-means clusters are right where
they're supposed to be: (1, 1), (1.2, 1), and (1.5, 1.3).

5. Do the macro evaluations measures provide good fits? Contrast with 2.

```{r}
evaluate(kmeans, dsd, measure = c("purity", "crand"), n = 500)
```

Our purity rating increased by 5% and our corrected Rand score is now
up from 27% to 84%, telling us that three macro-clusters is a much
better representation of the true structure of our data (as we should
expect). 






